Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-06-05 TEX/draft4.tex
Cosmetic
Some work on the 1D heterogeneous figure

In this paper, we argue that there is a gap between the two questions in the quoted excerpt above, and that one may be able to exploit it to design better reconstruction algorithms.
Specifically, the impossibility of particle picking does not necessarily imply impossibility of particle reconstruction.
%This can be understood from a statistical point of view by counting the number of parameters that one aims to estimate. \TODO{For a given target resolution, the number of parameters required to describe the 3-D model of the molecule is fixed: it does not depend on the amount of data collected. In contrast, particle picking involves estimation of several so-called nuisance parameters for each projection present in the micrographs, such as their locations for example. To overcome low $\SNR$s, the ratio of data to parameters must increase. This is impossible if the number of parameters grows with the amount of data collected, but not so if the number of parameters remains fixed. Simply put, the key is to bypass the particle picking stage.}
%%Thus, acquiring more data does not necessarily allow to improve the particle picking stage, whereas it should in principle allow to improve the molecule reconstruction.
%%Indeed, the signal itself (the many different views of the molecule) is still present in the micrograph: 
%Indeed, in principle at least, the micrograph does contain the necessary information (several views of the molecule(s) of interest). While we may not be able to locate these views, we may still be able to extract the relevant information through other means.
Indeed, the aim is only to reconstruct the molecule: estimating the locations of the particles in the micrograph is merely a helpful intermediate stage when it can be done. Our main message is that the limits particle packing imposes on molecule size do not translate into limits on particle reconstruction.

@@

In order to recover the planted image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelation functions of the micrographs to the autocorrelation functions of the planted image.
%To do so, we assume the image occurrences are well separated.
For any noise level, these autocorrelations can be estimated to any desired accuracy provided individual occurrences of the image are well separated and the image appears sufficiently many times in the micrographs. Importantly, there is no need to detect individual occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares or a phase retrieval algorithm; see Figure~\ref{fig:Einst_example} for an illustration and Section~\ref{sec:methods} for details. As a side note, we mention that expectation-maximization (EM)---a popular framework in SPR---is intractable for this problem; see Appendix~\ref{sec:theory} for a discussion.

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{vanheel1992correlation,shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required.

@@

\clearpage

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains four occurrences of the same $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

@@

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

>>>>

In this paper, we argue that there is a gap between the two questions in the quoted excerpt above, and that one may be able to exploit it to design better reconstruction algorithms.
Specifically, the impossibility of particle picking does not necessarily imply impossibility of particle reconstruction.
%This can be understood from a statistical point of view by counting the number of parameters that one aims to estimate. \TODO{For a given target resolution, the number of parameters required to describe the 3-D model of the molecule is fixed: it does not depend on the amount of data collected. In contrast, particle picking involves estimation of several so-called nuisance parameters for each projection present in the micrographs, such as their locations for example. To overcome low $\SNR$s, the ratio of data to parameters must increase. This is impossible if the number of parameters grows with the amount of data collected, but not so if the number of parameters remains fixed. Simply put, the key is to bypass the particle picking stage.}
%%Thus, acquiring more data does not necessarily allow to improve the particle picking stage, whereas it should in principle allow to improve the molecule reconstruction.
%%Indeed, the signal itself (the many different views of the molecule) is still present in the micrograph: 
%Indeed, in principle at least, the micrograph does contain the necessary information (several views of the molecule(s) of interest). While we may not be able to locate these views, we may still be able to extract the relevant information through other means.
Indeed, the aim is only to reconstruct the molecule: estimating the locations of the particles in the micrograph is merely a helpful intermediate stage when it can be done. Our main message is that the limits particle picking imposes on molecule size do not translate into limits on particle reconstruction.

@@

In order to recover the planted image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelation functions of the micrographs to the autocorrelation functions of the planted image.
%To do so, we assume the image occurrences are well separated.
For any noise level, these autocorrelations can be estimated to any desired accuracy, provided individual occurrences of the image are well separated and the image appears sufficiently many times in the micrographs. Importantly, there is no need to detect individual occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares or a phase retrieval algorithm; see Figure~\ref{fig:Einst_example} for an illustration and Section~\ref{sec:methods} for details. As a side note, we mention that expectation-maximization (EM)---a popular framework in SPR---is intractable for this problem; see Appendix~\ref{sec:theory} for a discussion.

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{vanheel1992correlation,shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required, thus significantly reducing concerns about model bias.

@@

%\clearpage

@@

\begin{figure}[t]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contain the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

@@

\TODO{Results: What answer was found to the research question; what did the study find? Was the tested hypothesis true?}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{heterogeneous_progressive_n12300000000_466300}
	\caption{An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

\TODO{Discussion: What might the answer imply and why does it matter? How does it fit in with what other researchers have found? What are the perspectives for future research?}
